A Tool for Visualizing Patterns of Spreadsheet Function Combinations \\ \large The Protracted and Generally Unpleasant Death of Justin A. Middleton\\ A Methodological Obituary

### 2016-03-15 doc/paper.tex
aaaaaaaaaaaaaaa paper

\section{Introduction}
% no \IEEEPARstart
The world of business owes a considerable debt to spreadsheets, the table-based interface which enables organization and manipulation of massive amounts of data. Their allure is in their versatility: while the end user does not need sophisticated programming skills to fill out a spreadsheet, the expert user has at their disposal numerous built-in operations, or functions, to expedite their work. As such, it should be of little surprise when Scaffidi and colleagues estimated that by 2012, over 50 million workers could be using them, including the 25 million who would be writing programs out of the functions included (citation and such).\par

Considering this ubiquity, it's crucial to get spreadsheets right. Something, something. \par

@@

This paper presents a tool, informed by such variegated sources, that visualizes not how people employ individual Excel functions but how they combine to make more sophisticated shreadsheet formulas.

@@

\section{Approach}
In visualizing the spreadsheet data, I aimed to accomplish the following goals:
\begin{itemize}
	\item The data should come from a varied enough sources to avoid having too narrow or specialized a scope.
	\item The tool should accommodate dataset sizes from the single spreadsheet to corpora of thousands.
	\item The tool should visualize as much data as possible in order to capture (and clearly indicate) both the most typical and the most anomalous cases...
	\item ...while being interactive enough to let the user choose which cases they see, as to not sacrifice readability.
	\item The tool should facilitate quantitative understanding by showing which functions and which combinations are used most frequently...
	\item ...and it should facilitate qualitative understanding by incorporating actual examples into the visualization from the datasets it studies, and it should help the user find these examples in their original context.
	\item The tool, by the same method as above, should thus guide a user from the abstract patterns of function usage into the specific and applied techniques of functions in their natural habitat.
\end{itemize} \par
Likewise, there were a few goals that I was specifically not trying to accomplish with this:
\begin{itemize}
	\item The tool should not be, by itself, responsible for generating functions from the data it collects. It is purely analytical.
	%TODO: I do have an idea which could actually do the below, but, eh, it would take
	%TODO: a fair amount of time to implement.
	\item The tool will not visualize isolated Excel functions, linking the specific instance to the abstract pattern. It works only the other way.
	\item Likewise, the tool will make no effort to explain what any given set of functions in combination actually does; the function of any given example is up to the user to infer.
\end{itemize}

\section{Case Study}
The patient exhibits strong cadaverous tendencies.

\section{Conclusion}
This paper is the end-all, be-all. There is no future work simply because the concept of future work cannot exist with reference to this paper. The work is done. I perfected it. Go home, kiss your love, kick back, call it a day. This is not just a good thing, but the best thing. Your job here is done, and I'm the one who did it.

>>>>

\section{Introduction}
% no \IEEEPARstart
Business and research alike owe their debts to spreadsheets, the table-based interface which empowers users to organize and manipulate huge bodies of data [citation for definition]. Their allure is in their versatility: while the novice end user can work without a deep knowledge of programming, the expert can expedite their work with a variety of built-in operations, or functions.\footnote{https://support.office.com/en-us/article/Excel-functions-alphabetical-b3944572-255d-4efb-bb96-c6d90033e188} As such, it should be of little surprise when Scaffidi and colleagues estimated that by 2012, over 50 million U.S. workers could be using them, including the 25 million who would be writing programs out of the functions included (Scaffidi citations).\par

Considering this ubiquity, it's crucial to get spreadsheets right. Our failures in their use can be ruinous, as in 2012 when the influential findings on economic growth were reversed by a selection error [Enron paper], among other horror stories.\footnote{http://www.eusprig.org/horror-stories.htm}

@@

This paper presents a tool, informed by such varied sources, that visualizes not how people employ individual Excel functions but how they combine to make more sophisticated spreadsheet formulas.

@@

\section{Approach}
\subsection{Goals}
In visualizing the spreadsheet data, I outlined a few core goals for what the tool should accomplish:
\begin{itemize}
	\item [1] \textbf{Draw an interactive interface to explore observed function combinations.} The combination space for all Excel functions is massive, let alone the space for observed formulas. Working from data with actual referents in practice, the tool must aid the user in navigating this space.
	\item [2] \textbf{Emphasize the quantitative patterns in formula construction.} The tool, accommodating datasets of a few spreadsheets to millions, should address the questions of how often the end users employed a certain function and where they used it. In this way, it must show precise metrics, such as frequency of  use and depth of function nesting, of the dataset it conveys.
	\item [3] \textbf{Promote a qualitative understanding of the patterns.} Lest these patterns remain abstract, the tool should supplement its observations with concrete instances of relevant formulas from the corpus. Ideally, it should even direct users to the exact cell in the spreadsheets where the formula was used, contextualizing the functions. 
\end{itemize} \par
Likewise, to bound our scope, we outlined a few goals to specifically avoid accomplishing with this:
\begin{itemize}
	\item [!1] \textbf{Do not try to directly explain what a function does.} Though the tool tries to foster tries understanding by linking pattern to example, it won't provide a precise description of what a function accomplishes. The tool's user must infer this.
	\item [!2] \textbf{Do not create new formulas.} This is essentially an exploratory tool, not a generative one. It should not produce any information other than new views of the original data. 
	%TODO: I do have an idea which could actually do the below, but, eh, it would take
	%TODO: a fair amount of time to implement.
	\item [!3] \textbf{Do not visualize individual formulas.} Though there is room to explore the place of a single function in the group, the tool must design the core visualization around the entire body of data, not the other way around.
\end{itemize}

\subsection{Design}
Given the composition of formulas as functions and their arguments (which could be yet more functions), we need a visualization style that conveys this sense of parent-child relationship. As such, we decided to represent formulas as trees where the branching factor depends on both the number of arguments in a function and the number of possible functions observed as an argument in a function. \par
Guided by these goals, we faced a number of decision points in our design, which we outline below.

\begin{itemize}
	\item [1a] \textit{Copied formulas}: Excel allows users to spread a formula over an area, repeating the same task in each cell with minor adjustments. Without checking for this, the analysis may not reveal the functions most commonly used together but rather the formulas most often applied to large areas. To combat this, we converted formulas from their native A1 format to the relative R1C1, in which copied formulas should be identical, and reduced the records to unique R1C1 formulas per sheet.
	\item [1b] \textit{Optional arguments}: How should the tool handle functions which accept a variable number of arguments? SUM, for example, can have anywhere from 1 to 255, and IF can accept either 2 or 3. \footnote{https://support.office.com/en-us/article/Excel-functions-by-category-5f91f4e9-7b42-46d2-9bd1-63f26a86c0eb} Without prior evidence, it's possible that spreadsheet programmers use different techniques for different numbers arguments. To account for this, we separate and analyze the different quantities of arguments observed for each function; the tool, however, uses as default all of the options collapsed into a single representation.
\end{itemize}

\begin{itemize}
	%TODO: A picture would go well here.
	\item [2a] \textit{Importance of depth}: When a function appears within another, should it be analyzed only as a nested function or would it also be valid to analyze the nested function on its own? If the former, then information about the same function will be scattered across different trees with no way to aggregate them. If the latter, then the tool will analyze some functions multiple times to capture every possible level of nesting. Both approaches have benefits and drawbacks, and so we included both.
	\item [2b] \textit{Pattern density}: How should the tool quantitatively order its elements: by a function's raw frequency or by the unity of patterns it leads to? For example, if SUM has for its first argument two possibilities, one which itself contains 1000 unique argument possibilities with 1 occurrence each (high frequency, low pattern density) and another seen with 2 argument possibilities of 100 occurrences each (low frequency, high pattern density), which would be more interesting to emphasize? The answer depends on the nuances of the questions, but for simplicity, I've shown the former.
	\item [2c] \textit{Non-functions}: How should the tool represent everything in the formula that isn't a function: numbers, string literals, errors, references, etc? Since these don't accept arguments, they will adorn the tree as leaves, and their precise content won't affect the functions around them as long as their types are known. As such, in the visualization, all of these nodes are replaced and aggregated under their types.
\end{itemize}

\begin{itemize}
	\item [3a] \textit{Suitable examples}: Goal 3 supports the inclusion of examples in the visualization as a way of tying pattern to example, but how should examples be chosen? For this, the simpler is the better, and we made the broad working assumption that shorter (by character length) functions are simpler, and thus chose the shortest available.
	\item [3b] \textit{Spreadsheet connections}: Is it possible to contextualize these examples even better? Yes, by leading the user directly back to the originating spreadsheet. The tool hyperlinks each example, then, back to the file that contains it, providing sheet, row, and column numbers if it doesn't open directly there.
\end{itemize}

\subsection{Implementation}
We can view the final visualization as the product of two discrete processes:
\begin{itemize}
	\item \textit{Collection}: Given a set of Excel sheets, the tool, written mostly in Java, uses Apache POI\footnote{https://poi.apache.org/} to identify and iterate over every cell containing a valid formula. Afterward, it calls POI's formula parser to break the formula text into an ordered set of individual tokens, which the tool then parses into the tree-like form by which it is recorded. When all formulas have been analyzed like this, it produces JSON files for each top-level function in the set.
	\item \textit{Presentation}: The JSON files, meanwhile, feed into the presentation code, implemented in Javascript with much help from the visualization library D3\footnote{https://d3js.org/}. 
\end{itemize}
When the presentation code displays the tree, it shows two types of nodes with different meanings: the circles represent a function that has been observed at that structural place, and hovering over these yields a tooltip with supplementary information and examples; and squares, which represent the argument positions for the function which is their parent. For example, if the IF function has been observed with three arguments, 3 squares will extend from it as children. Clicking on the square labeled '1' will yield more circles, all of which are functions or values that the tool observed as the first argument in an IF function; clicking on '2' will yield those possibilities in the second position; and so on.

\subsection{Limitations}
Because the data collection depends on POI's formula parser, it affords no leeway or partial information from a formula; it either processes it perfectly or throws it out. As such, the visualization inhibits insight into anything with syntactical errors. Note, however, that this does not include standard Excel errors like \#REF! and \#DIV/0, which the parser handles well. \par
Unfortunately, the parse also failed to handle any externally defined functions, even if they were syntactically correct in their original context. On one hand, we don't consider the loss of these specific functions to be tragic, as they are not part of Excel proper and thus provide no insight into how people use built-in functions. However, information about other functions within them are lost. \par
Many functions have an expected order of arguments, but some, like SUM and MATCH, can be reordered in various ways and maintain their value. However, the representation that this tool uses reinforces for all functions that the order is important, even in these exceptions. \par

\section{Case Study}
---

\section{Conclusion}
---
